---
title: "Exercise 4.2: Exploring the pertussis dataset"
author: "Lieven Clement, and Milan Malfait"
date: "statOmics, Ghent University (https://statomics.github.io)"
---

Researchers wanted to study the immune response on pertussis. 
They have set up an experiments with 40 rats. 
16 rats were infected with pertussis and 24 rats received a control treatment. 
Researchers measured the white blood cell concentration (WBC) in each rat (count per mm$^3$. 

De data consists of two variables:

- WBC: white blood cell count (counts/mm$^3$). 
- trt: treatment

    - control: rat recieved control treatment
    - pertussis: rat was infected with pertussis

Load the libraries

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

# Import the dataset

Data path:

  `https://raw.githubusercontent.com/statOmics/PSLS21/data/wbcon.csv`

```{r, eval=FALSE}
wbcon <-
```

```{r, eval=FALSE}
glimpse(wbcon)
```

# Goal

The overarching goal of this study was to assess if the white blood cell count is 
transplanting by pertussis infection. To this end, the researchers randomized 40 rats
to two treatments: A control treatment and a treatment with pertussis infection.

In the tutorial on hypotheses testing we will use a formal statistical test to assess if the white blood cell count is on average different between infected and control rats. 

For this test to be valid, we have to assess following assumptions:

1. The data in each treatment group is normally distributed.

2. The data from the two treatment groups has the same variance.

A statistical analysis always starts with a exploratory data analysis to get
insight in the nature and distribution of the data, and to assess the
assumptions of the downstream data analysis. Mastering this data exploration
step is the purpose of the current exercise.

# Data visualization

A crucial first step in a data analysis is to visualize and to explore the raw data.

## Histogram 

First, make a histogram of the data. Fill in the
missing parts in the chunk of code below:

```{r, eval=FALSE}
wbcon %>%
  ggplot(aes(x=...,fill=...)) + ## fill in the correct values for x and fill
  geom_histogram() +
  facet_grid(rows = vars(...)) + ## fill in to put the histograms for both treatment conditions in a separate row
  theme_bw() +
  xlab("relative abundance (%)")
```

Based on this plot, it seems that the white blood cell count 
is ... for rats that were randomized to the pertussis infection than rats that received the control treatment.

## Boxplot 

However, given the small sample size a better option to visualize these data are `boxplots`:

```{r, eval=FALSE}
wbcon %>%  
  ggplot(aes(x=...,y=...,fill=...)) +
  
```

What do you observe?

## QQ-plots 

To assess the assumption that the data are normally distributed in each treatment group, we will use QQ plots.

```{r, eval=FALSE}
wbcon %>%  
  ggplot(aes(sample=...)) +
  ... +
  ... +
  facet_grid(...)
```

What do you observe? 

# Descriptive statistics

Here, we will generate some informative descriptive statistics
for the dataset.

We first summarize the data and calculate the mean, standard
deviation, number of observations and standard error and store the
result in an object wbcSum via 'wbcSum<-`

1. We pipe the `wbcon` dataframe to the group_by function to group
the data by treatment groep `group_by(groep)`
2. We pipe the result to the `summarize()` function to summarize
the "WBC" variable and calculate the mean, standard deviation and
the number of observations
3. We pipe the result to the `mutate` function to make a new
variable in the data frame that is named `se` for which we calculate the
standard error ($\sigma / n$)

```{r, eval=FALSE}
## Use the instructions from above to generate the summary statistics
...
```

This concludes the data exploration. In the next tutorial sessions, we will learn how to formally
test if the observed difference in WBC between rats that were infected with pertussis and those receiving the control treatment is **statistically significant**.
