---
title: "Exercise 4.2: Exploring the armpit dataset"
author: "Lieven Clement and Jeroen Gilis"
date: "statOmics, Ghent University (https://statomics.github.io)"
output:
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

In this exercise, we will explore the "smelly armpit" dataset.

# Smelly armpit dataset

Smelly armpits are not caused by sweat, itself. The smell is caused
by specific micro-organisms belonging to the group of
*Corynebacterium spp.* that metabolise sweat. Another group of abundant bacteria
are the *Staphylococcus spp.*, these bacteria do not metabolise sweat in smelly
compounds.

The CMET-group at Ghent University does research to on transplanting the armpit
microbiome to save people with smelly armpits.

- Proposed Therapy:
  	1. Remove armpit-microbiome with antibiotics
    2. Influence armpit microbiome with microbial transplant, see this 2 minute
       talk on [youtube](https://youtu.be/9RIFyqLXdVw)

- Experiment:

    - 20 students with smelly armpits are attributed to one of
      two treatment groups
    - placebo (only antibiotics)
    - transplant (antibiotica followed by microbial transplant).
    - The microbiome is sampled 6 weeks upon the treatment
    - The relative abundance of *Staphylococcus spp.* on
      *Corynebacterium spp.* + *Staphylococcus spp.* in the
      microbiome is measured via DGGE (*Denaturing Gradient Gel
      Electrophoresis*).

Load the libraries

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

# Import the dataset

```{r, eval=FALSE}
ap <-
```

```{r, eval=FALSE}
glimpse(ap)
```

# Goal

The overarching goal of this study was to assess if the relative abundance
*Staphylococcus spp.* in the microbiome of the armpit is affected by
transplanting the microbiome. To this end, the researchers randomized patients
to two treatments: A treatment with antibiotics only and a treatment with
antibiotics and a microbial transplant.

In the tutorial on hypotheses testing we will use a formal statistical test to
generalize the results from the sample to that of the population.
For this test to be valid, we have to assess following assumptions:

1. The data in each treatment group is normally distributed.

2. The data from the two treatment groups has the same variance.

A statistical analysis always starts with a exploratory data analysis to get
insight in the nature and distribution of the data, and to assess the
assumptions of the downstream data analysis. Mastering this data exploration
step is the purpose of the current exercise

# Data visualization

A crucial first step in a data analysis is to visualize and to explore the raw
data.

First, try to make a histogram of the data. Fill in the
missing parts in the chunk of code below to get a good-looking
visualization:

```{r, eval=FALSE}
ap %>%
  ggplot(aes(x=...,fill=...)) + ## fill in the correct values for x and fill
  geom_histogram() +
  facet_grid(rows = vars(...)) + ## fill in to put the histograms for both treatment conditions in a separate row
  theme_bw() +
  xlab("relative abundance (%)")
```

Based on this plot, it seems that the relative abundance
is higher for subjects who had the transplants.
However, given the small sample size the histograms
are not optimally informative. A better option for these data would be to
show the data in a `boxplot`:

```{r, eval=FALSE}
ap %>%  ggplot(aes(x=...,y=...,fill=...)) +
```

What do you observe?

# Descriptive statistics

Here, we will generate some informative descriptive statistics
for the dataset.

We first summarize the data and calculate the mean, standard
deviation, number of observations and standard error and store the
result in an object apRelSum via 'apRelSum<-`

1. We pipe the `ap` dataframe to the group_by function to group
the data by treatment trt `group_by(trt)`
2. We pipe the result to the `summarize_at` function to summarize
the "rel" variable and calculate the mean, standard deviation and
the number of observations
3. We pipe the result to the `mutate` function to make a new
variable in the data frame that is named `se` for which we calculate the
standard error

```{r, eval=FALSE}
## Use the instructions from above to generate the summary statistics
...
```

This concludes the data exploration. Tomorrow, we will learn how to formally
test if the observed difference is **statistically significant**.
