---
title: "Exercise 4.3: Exploring the captopril dataset"
author: "Lieven Clement, Jeroen Gilis and Milan Malfait"
date: "statOmics, Ghent University (https://statomics.github.io)"
---

# The captopril dataset

The `captopril dataset` holds information on a small experiment with
15 patients that have increased blood pressure values.
More specifically, for each patient we will have four values;
one value for systolic blood pressure and one for diastolyic,
both before and after treating the patient with a drug named captopril.

# Import the data

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(readr)
library(ggplot2)
```

```{r}
captopril <- read_csv("https://raw.githubusercontent.com/statOmics/PSLS21/data/captopril.txt")
```

```{r}
head(captopril)
```

**Why is this dataset not tidy?**

# Data visualization

Let's say we now first want to visualize the data.
One possibility to easily visualize the four types
of blood pressure values is by adopting the `gather`
function from tidyverse. It will reshape the dataframe,
such that we have have a single variable `type`, which
points at one of the four blood pressure types, and `bp`,
which points at the actual value for each type
for each patient. We call this data format also a *long*
format, while the original dataset is in *wide* format.

**Side note**: `gather()` is a rather outdated function from the tidyverse (though it still works
perfectly fine!), the current recommended approach is to use
[`pivot_longer()`](https://tidyr.tidyverse.org/reference/pivot_longer.html)

```{r}
## Pivot data from wide to long format
captopril_long <- captopril %>%
  gather(type, bp, -id)

## Print the result
captopril_long
```

We transformed the data and stored it in a new object `captopril_long` for easy reuse later.  To
improve the visualizations below, we can also recode the `type` variable so it's a bit more
descriptive; `SBPb` correspond to "Systolic Blood Pressure before treatment", `SBPa` to "SBP after
treatment" and so on. We can use another package from the tidyverse named
[*stringr*](https://stringr.tidyverse.org) to transform the character `type` variable.

```{r}
captopril_long <- captopril_long %>%
  mutate(
    type = str_replace(type, "b", "_before"),
    type = str_replace(type, "a", "_after")
  )

## Print the result
captopril_long
```

Finally, R will by default visualize any character variables in alphabetic order. However, in this case, it makes more sense to put the `before` types first. We can do this by transforming the characgter variable `type` to a factor and use (yet) another tidyverse package (there are about 8 of them!) called [*forcats*](https://forcats.tidyverse.org/) to make the `before` variables the reference level (i.e. the first in order).

```{r}
captopril_long <- captopril_long %>%
  mutate(
    type = fct_relevel(
      type,
      c("SBP_before", "SBP_after", "DBP_before", "DBP_after")
  ))
captopril_long
```

*Note that this doesn't change anything to the order in which the values appear in the data
(compare the output with that from before. The only thing that changed here is that we converted
`type` to a __factor__, indicated by `<fct>` when printing the `tibble` object.*

## Barplot

Not all visualization types will be equally informative.
Let us first make a barplot of the data. A barplot is a plot that you will
commonly find in scientific publication The code for generating such a barplot
is provided below:

```{r}
captopril_long %>%
  ## Calculate summarry statistics for the "bp" variable for each "type"
  group_by(type) %>%
  summarize(
    mean = mean(bp, na.rm = TRUE),
    sd = sd(bp, na.rm = TRUE),
    n = n()
  ) %>%
  ## Compute the standard errors for the means
  mutate(se = sd / sqrt(n)) %>%
  ggplot(aes(x = type, y = mean, fill = type)) +
  scale_fill_brewer(palette = "RdGy") +
  theme_bw() +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  ggtitle("Barplot of different blood pressure measurements") +
  ylab("blood pressure (mmHg)")
```

A barplot, however, is not very informative. The height of the
bars only provides us with information of the mean blood pressure.
However, we don't see the actual underlying values, so we for
instance don't have any information on the spread of the data.
It is usually more informative to represent to underlying
values as _raw_ as possible. Note that it is possible to add the
raw data on the barplot, but we still would not see any measures
of the spread, such as the interquartile range.

**Based on this critisism, can you think of a better**
**visualization strategy for the captopril data?**

**Add your proposed visualization strategy here**

## Boxplot

It is usually more informative to represent to underlying
values as _raw_ as possible. Boxplots are ideal for this!

Note that it in theory it is possible to add the raw data on
the barplot, but we still would not see any measures of the
spread, such as the interquartile range.

```{r}
captopril_long %>%
  ggplot(aes(x = type, y = bp, fill = type)) +
  scale_fill_brewer(palette = "RdGy") +
  theme_bw() +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2) +
  ggtitle("Boxplot of different blood pressure measurements") +
  ylab("blood pressure (mmHg)") +
  stat_summary(
    fun = mean, geom = "point",
    shape = 5, size = 3,
    color = "black"
  )
```

With the boxplot, we get a lot of useful information.
We immediately see multiple features of the spread of
the data, such as the median, the 25% and 75% quantiles
and outliers. Since we only have 15 raw values (patients),
we can easily add them to the plot without getting messy.

In terms of interpretation, we can see that the median
systolic and diastolic blood pressure values are lower
after treatment with captopril than before. If we want to
have visual inference on the mean values (cfr. class on t-test),
we can add them to the plot `stat_summary` function.

## Scatterplot

An important feature of this dataset is that it contains
paired data; for each patient, we have blood pressure values
(systolic and diastolic) before and after treatment with
captopril.

We can visualize this as follows;

```{r}
captopril_long %>%
  filter(type %in% c("SBP_after", "SBP_before")) %>%
  ggplot(aes(x = id, y = bp, color = type)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("blue", "red")) +
  theme_bw()
```

Or, alternatively, by creating a `line plot`:

```{r}
captopril_long %>%
  filter(type %in% c("SBP_after", "SBP_before")) %>%
  ggplot(aes(x = type, y = bp)) +
  geom_line(aes(group = id)) +
  theme_bw()
```

We see that for all patients, the systolic blood
pressure is lower after captopril treatment than before.

Note that we could not see this from the boxplot directly;

```{r}
captopril_long %>%
  filter(type %in% c("SBP_after", "SBP_before")) %>%
  ggplot(aes(x = type, y = bp, fill = type)) +
  scale_fill_brewer(palette = "RdGy") +
  theme_bw() +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2) +
  ggtitle("Boxplot of blood pressure measurements before and after treatment") +
  ylab("blood pressure (mmHg)") +
  stat_summary(fun = mean, geom = "point", shape = 5, size = 3, color = "black", fill = "black")
```

A typical next step is to perform a test to find out
whether the mean systolic blood pressure value after
captopril treatment is _significantly_ different from
the mean systolic blood pressure value before treatment.

Analagously, we may subtract the `after` measurement from the
`before` measurement, and test if the difference between the
two sets of values is significantly different from zero.

```{r}
captopril %>%
  mutate(bp_diff = SBPa - SBPb) %>%
  select(bp_diff) %>%
  ggplot(aes(x = "", y = bp_diff)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2) +
  ggtitle("Boxplot of the difference in blood pressure") +
  ylab("blood pressure difference (mmHg)") +
  stat_summary(
    fun = mean, geom = "point",
    shape = 5, size = 3, color = "black") +
  theme_bw() +
  ylim(-40, 10) +
  geom_hline(yintercept = 0, color = "red") ## adds horizontal line to plot
```

The mean difference is lower than zero. It seems that on average the captopril
treatment has lowered the blood pressure by about 20 mmHg. Tomorrow, we will
show how we can test if this reduction is significantly different from 0 mmHg.
