---
title: "Exercise 4.3: Exploring the captopril dataset"
author: "Lieven Clement and Jeroen Gilis"
date: "statOmics, Ghent University (https://statomics.github.io)"
output:
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

# The captopril dataset

The `captopril dataset` holds information on a small experiment with
15 patients that have increased blood pressure values.
More specifically, for each patient we will have four values;
one value for systolic blood pressure and one for diastolyic,
both before and after treating the patient with a drug named captopril.

# Import the data

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(readr)
library(ggplot2)
```

```{r}

```

Have a first look at the data

```{r}

```

# Data visualization

Let's say we now first want to visualize the data.
One possibility to easily visualize the four types
of blood pressure values is by adopting the `gather`
function from tidyverse. It will reshape the dataframe,
such that we have have a single variable `type`, which
points at one of the four blood pressure types, and `bp`,
which points at the actual value for each type
for each patient.

```{r, eval=FALSE}
captopril %>%
  gather(type,bp,-id)
```

## Barplot

Not all visualization types will be equally informative.
Let us first make a barplot of the data. A barplot is a plot that you will
commonly find in scientific publication The code for generating such a barplot
is provided below:

```{r, eval=FALSE}
captopril %>%
  gather(type,bp,-id) %>%
  group_by(type) %>%
     summarize_at("bp",
               list(mean=~mean(.,na.rm=TRUE),
                    sd=~sd(.,na.rm=TRUE),
                    n=function(x) x%>%is.na%>%`!`%>%sum)) %>%
  mutate(se=sd/sqrt(n)) %>%
  ggplot(aes(x=type,y=mean,fill=type)) +
    scale_fill_brewer(palette="RdGy") +
    theme_bw() +
    geom_bar(stat="identity") +
    geom_errorbar(aes(ymin=mean-se, ymax=mean+se),width=.2) +
    ggtitle("Barplot of different blood pressure measures") +
    ylab("blood pressure (mmHg)")
```

A barplot, however, is not very informative. The height of the
bars only provides us with information of the mean blood pressure.
However, we don't see the actual underlying values, so we for
instance don't have any information on the spread of the data.
It is usually more informative to represent to underlying
values as _raw_ as possible. Note that it is possible to add the
raw data on the barplot, but we still would not see any measures
of the spread, such as the interquartile range.

**Based on this critisism, can you think of a better**
**visualization strategy for the captopril data?**

**Add your proposed visualization strategy here**

```{r, eval=FALSE}

```

# Descriptive statistics

- Generate a code chunk to calculate useful summary statistics for
the capropril data
