---
title: "Exercise 4.3: Exploring the diabetes dataset"
author: "Lieven Clement, Jeroen Gilis and Milan Malfait"
date: "statOmics, Ghent University (https://statomics.github.io)"
---

# Aims of this exercise

In this exercise, you will acquire the skills to 

- recognize paired data
- conduct a data exploration in R for data from
paired experimental designs.
- interpret the results of a data exploration for paired experimental designs

# The diabetes dataset

The `diabetes dataset` holds information on a small experiment with
8 patients that are subjected to  a glucose tolerance test.

Patients had to fast for eight hours before the test. 
When the patients entered the hospital their baseline glucose level was measured (mmol/l). 

Patients then  had to drink 250 ml of a syrupy glucose solution containing 100 grams of sugar. 
Two hours later, their blood glucose level was measured again. 

The data consist of three variables:

- before: glucose concentration upon 8 hours of fasting (mmol/l)
- after: glucose concentration 2 hours after drinking glucose solution (mmol/l).
- patient: identifier for the patient

# Import the data

Data path:

  `https://raw.githubusercontent.com/statOmics/PSLS21/data/diabetes.txt`

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(readr)
library(ggplot2)
```

```{r, eval = FALSE}
diabetes <- ...
```

Have a first look at the data

```{r}

```

# Data visualization

The data are not in tidy format. 
We can easily visualize the glucose levels by adopting the `gather()`
function from tidyverse. 
It will reshape the dataframe,
such that we have have a single variable `time`, which
points at the timepoint of the measurement, and `concentration`,
which contains the actual glucose concentration at before or after drinking the glucose solution
for each patient.

```{r, eval=FALSE}
diabetes %>%
  gather(time, concentration, -patient)
```

## Barplot

Not all visualization types will be equally informative.

A barplot is a plot that you will
commonly find in scientific publications. 
The code for generating such a barplot
is provided below:

```{r, eval=FALSE}
 diabetes %>%
  gather(time, concentration, -patient) %>% 
  ## Calculate summarry statistics for the "bp" variable for each "type"
  group_by(time) %>%
  summarize(
    mean = mean(concentration, na.rm = TRUE),
    sd = sd(concentration, na.rm = TRUE),
    n = n()
  ) %>%
  ## Compute the standard errors for the means
  mutate(se = sd / sqrt(n)) %>%
  ggplot(aes(x = time, y = mean, fill = time)) +
  theme_bw() +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  ggtitle("Barplot of glucose measurements") +
  ylab("concentration (mmol/l)")
```


A barplot, however, is not very informative. 
The height of the
bars only provides us with information of the mean blood pressure.
However, we don't see the actual underlying values, so we for
instance don't have any information on the spread of the data.
It is usually more informative to represent to underlying
values as _raw_ as possible. 
Note that it is possible to add the
raw data on the barplot, but we still would not see any measures
of the spread, such as the interquartile range.

Another crucial aspect of the data are also not displayed: 
the data are paired!

**Based on these critisisms, can you think of a better**
**visualization strategy for the captopril data?**

**Add your proposed visualization strategy here**

```{r, eval=FALSE}

```

# Descriptive statistics

- Generate a code chunk to calculate useful summary statistics for
the diabetes data
